% Conclusion: direct answer to Project 7's question

\section{Discussion: how hard is the problem?}\label{sec:conclusion}

The Project~7 prompt asks: \emph{``How hard is [cardinality-constrained portfolio optimisation
on real data] from an empirical computational point of view?''} Our answer is a calibration:
\textbf{the difficulty depends on which of three regimes the instance lives in}, and on the
real S\&P~500 panel the data is \emph{not} in the hardest regime.

\subsection{Three regimes of difficulty}\label{ssec:regimes}

Parameterise the instance by the planted-spike SNR $a := \alpha/\sqrt n$ as in
\S\ref{sec:ogp}. Theorems~\ref{thm:detect} and \ref{thm:forbid} together with Proposition~\ref{prop:slh-recovery}
identify three regimes:

\medskip
\noindent\textbf{Regime I (easy and undetectable), $a < a_\star(\rho)$.} Below the detection
threshold the planted signal is statistically indistinguishable from random fluctuation:
no algorithm (information-theoretically) can identify $\sigma^\star$. From a practical
perspective this regime is uninteresting, since the maximiser of $J$ on a random instance is
just a random subset with no exploitable structure.

\smallskip
\noindent\textbf{Regime II (OGP-hard), $a_\star(\rho) < a < a_{\mathrm{rec}}$.} Detection is
possible but \emph{exact recovery} is blocked for path-stable algorithms: any algorithm whose
support is $\eta$-Hamming-stable along a Gaussian bridge fails to interpolate from a random
seed to the planted optimum without leaving the near-optimal level set
(Corollary~\ref{cor:hardness}). LASSO, top-eigenvector, and constrained Metropolis at fixed
$\beta$ all live in this class and all fall victim to the OGP plateau. On synthetic data
calibrated to this regime we observe exactly this barrier (Figure~\ref{fig:synth-overlap}).

\smallskip
\noindent\textbf{Regime III (easy and recoverable), $a > a_{\mathrm{rec}}$.} Above the
Bonferroni recovery threshold $a_{\mathrm{rec}}=\nu\sqrt\rho(\sqrt{2\log(n-k)}+\sqrt{2\log k})$,
even the simplest non-stable algorithm (top-$k$ thresholding of $\hat\mu$, equivalent to
\textsc{Slh}'s spectral seed) recovers $\sigma^\star$ with high probability
(Proposition~\ref{prop:slh-recovery}).

\subsection{Where does the real data live?}\label{ssec:realwhere}

We can read the empirical SNR off the Phase~2a results: the brute-force oracle $\sigma^\star$
recovers a return $\hat\mu^\top\sigma^\star/k$ that is approximately $0.6$ to $1.2$ standard
deviations of the bulk mean above the median asset. In our parametrisation, after
factor-residualising $\hat\mu$ to obtain the idiosyncratic Sharpe component, the empirical
$a$ on the $(n,k)=(50,5)$ instance is approximately $a_{\mathrm{emp}}\approx 1.5$ to $2.0$, sitting
\emph{just above} $a_\star(\rho=0.10)\approx 0.90$ but \emph{below} the recovery threshold
$a_{\mathrm{rec}}\approx 2.6$. Empirically this places the real data \emph{at the boundary
of Regime~II and Regime~III}.

The boundary placement is exactly what the Phase~2a numbers reveal: simulated annealing
(unstable, OGP-immune) hits the oracle on every cell within the brute-force range; LASSO
(stable, OGP-vulnerable) is uniformly off by $4$ to $11$\,bp; top-eigenvector (also stable)
is off by $11$ to $22$\,bp; and \textsc{Slh}\ (unstable by design, with the polish stage doing
the recovery work the seed cannot) achieves $0$\,bp on every cell.

\subsection{Direct answer to the question}\label{ssec:answer}

\begin{itemize}\itemsep=2pt
\item \textbf{In the empirical regime, the problem is computationally tractable.} A
deterministic polynomial-time algorithm (\textsc{Slh}, $0.07$ to $0.13$\,ms per call on the
$(n,k)$ grid we test) matches the brute-force oracle on $100\%$ of the cells.
\item \textbf{The reason is that the empirical SNR sits at the boundary of Regimes II and III},
not deep inside the OGP-hard region. The \emph{statistical} content of the problem makes
exact recovery information-theoretically possible, and the discrete combinatorial structure
of \textsc{Slh}'s top-$k$ thresholding plus 1-swap polish is sufficient to extract it.
\item \textbf{If real data ever moved deeper into Regime~II} (lower SNR per asset, larger $n/k$),
no convex-relaxation or stable iterative method would work. The OGP theory is then a
\emph{prediction of the practitioner experience} that ``LASSO doesn't help once the signal
gets weak'', a folklore observation given a rigorous probabilistic basis here.
\item \textbf{The bottleneck is statistical, not computational.} On real data even the
oracle's out-of-sample Sharpe (test-block) drops by $30$ to $50\%$ relative to its in-sample
Sharpe. Optimising harder over the in-sample objective doesn't buy out-of-sample performance;
this is the financial counterpart to \emph{generalisation in high-dimensional learning}.
\end{itemize}

\subsection{Open theoretical questions}\label{ssec:open}

\begin{enumerate}\itemsep=1pt
\item \emph{Classical (interior) OGP for the full quadratic Hamiltonian.}
\S\ref{ssec:ogp-discussion} conjectures (and Phase~3b empirically gestures at, but does
not prove) an interior forbidden interval $(y_\ell,y_h)\subsetneq(\rho,1)$ for $J$ at
intermediate $\beta$. The conjecture would follow from a Sherrington-Kirkpatrick-style
second-moment analysis of the quadratic-form spike-plus-bulk model; the analysis is
non-trivial because $\hat\Sigma$ is not isotropic.
\item \emph{Sharpening Proposition~\ref{prop:slh-recovery}.} The Bonferroni bound
$a_{\mathrm{rec}}$ is conservative by a $\Theta(\log\log n / \sqrt{\log n})$ factor (Phase~4a
empirical curve crosses 50\% at $a\approx 2.25$ vs.\ predicted $2.64$). Tightening to the
sharp extreme-value asymptotic for the maximum of $n$ i.i.d.\ Gaussians would close the gap.
\item \emph{Out-of-sample generalisation.} The optimisation pipeline can be combined with a
shrinkage estimator of $\hat\mu, \hat\Sigma$ (e.g., Ledoit-Wolf) to control the
overfitting documented in \S\ref{ssec:p2a-results}. Whether this changes the OGP regime is
a non-trivial question, since shrinkage shifts the effective SNR, possibly across regime
boundaries.
\end{enumerate}

\subsection*{Code and reproducibility}

All experiments are reproducible from the notebooks
\texttt{extract\_data.ipynb}, \texttt{01\_diagnostics.ipynb},
\texttt{02\_oracle\_baselines.ipynb}, \texttt{03\_mcmc\_sampler.ipynb},
\texttt{04a\_ogp\_theory.ipynb}, \texttt{04\_ogp\_experiment.ipynb},
\texttt{05\_hybrid\_algorithm.ipynb}. Total wall-clock to regenerate every figure and
table in this writeup: under five minutes on a single laptop CPU, with deterministic
seeds set throughout.
