\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathrsfs}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{cleveref}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}

\title{Cardinality-Constrained Portfolio Optimization \\
  via the Overlap Gap Property: \\
  An Application of Discrete Probability and Stochastic Processes}
\author{Auritro Dutta, Aleksandr Marteev, Aviral Aggarwal, Alan Berdikulov \\
  MIT 6.7720 / 15.070 / 18.619, Discrete Probability and Stochastic Processes}
\date{Spring 2026}

\begin{document}
\maketitle

\begin{abstract}
We study cardinality-constrained portfolio selection on $n=468$ S\&P~500 constituents
over a 10-year window ($T=2{,}512$ trading days), asking \emph{how hard is this problem
empirically?} Two answers, in opposite directions:

\textbf{Computationally, the in-sample problem is easy.} A deterministic spectral seed
plus 1-swap polish (\textsc{Slh}) matches the brute-force oracle on every $(n,k)$ cell we
can verify by enumeration, in $0.1$\,ms per call. The empirical SNR sits at the boundary
of two informational regimes: the detection regime, where the planted optimum is
identifiable, and the recovery regime, where simple top-$k$ thresholding suffices.

\textbf{Statistically, the problem is hard.} The brute-force oracle's out-of-sample
Sharpe falls by $30$--$50\%$ relative to in-sample. Greedy, simulated annealing, and
\textsc{Slh} all inherit the same overfitting, because they all recover the same
support. The bottleneck is not computation; it is estimation of $\hat\mu,\hat\Sigma$.

For the planted linear score model $H(\sigma)=\hat\mu^\top\sigma$ we prove a first-moment
detection upper bound $a_\star(\rho)= (1-\rho)^{-1}\sqrt{2 h(\rho)}$
(\cref{thm:detect}) and a single-sided forbidden-overlap interval
(\cref{thm:forbid}), both verified by constrained Metropolis. For the full quadratic
Hamiltonian $J(\sigma)=\tfrac{1}{k}\hat\mu^\top\sigma-\tfrac{\gamma}{2k^2}\sigma^\top
\hat\Sigma\sigma$ we present empirical evidence consistent with a classical interior
OGP, namely a sharp mixing-time blow-up of constrained Metropolis on $(n,k)=(40,10)$;
a rigorous proof remains open.
\end{abstract}

\section{Introduction}\label{sec:intro}

\paragraph{The problem.} Given $n$ stocks with daily returns $R\in\mathbb R^{T\times n}$,
sample mean $\hat\mu = \tfrac1T R^\top\mathbf 1$ and sample covariance $\hat\Sigma$, fix a
target portfolio size $k$ and risk-aversion $\gamma>0$. Following Markowitz \cite{markowitz},
the cardinality-constrained mean-variance problem is
\begin{equation}\label{eq:problem-intro}
  \sigma^\star \;=\; \arg\max_{\sigma\in\Omega_{n,k}} J(\sigma),
  \qquad J(\sigma) := \frac{1}{k}\,\hat\mu^\top\sigma - \frac{\gamma}{2k^2}\sigma^\top\hat\Sigma\sigma,
\end{equation}
with $\Omega_{n,k}=\{\sigma\in\{0,1\}^n:\sum_i\sigma_i=k\}$. The feasible set has size
$\binom{n}{k}$; brute force is out past $\sim\!10^7$, e.g.\ $\binom{50}{10}\approx 10^{10}$.

\paragraph{Two questions.} We attack the problem with two questions in mind.
\begin{enumerate}\itemsep=2pt
\item \textbf{In-sample}: how hard is it to solve \eqref{eq:problem-intro} on the empirical
$(\hat\mu,\hat\Sigma)$? This is a discrete optimisation problem on a specific noisy input.
\item \textbf{Out-of-sample}: does the answer to (1) translate into a portfolio that
actually performs on held-out data? This is a generalisation problem.
\end{enumerate}
Our headline finding is that the two answers diverge sharply. In-sample, the problem is
computationally easy on real S\&P~500 data: a deterministic algorithm matches the
brute-force oracle in $0.1$\,ms per call. Out-of-sample, even the oracle's Sharpe drops
$30$--$50\%$ relative to in-sample. The bottleneck is not algorithmic; it is the noise
in $(\hat\mu,\hat\Sigma)$ itself.

\paragraph{Why discrete probability?} The interplay between empirical
$(\hat\mu,\hat\Sigma)$ (a sample-noise object) and the combinatorial structure of
$\Omega_{n,k}$ puts the problem in the territory of probabilistic combinatorics. We use:
(i) concentration for $\hat\mu,\hat\Sigma$, controlling the strength of in-sample
rankings; (ii) the spectrum of $\hat\Sigma$ via random-matrix theory, controlling whether
convex relaxations recover the combinatorial optimum; and (iii) the Overlap Gap Property
(OGP) of \cite{gamarnik-zadik-clique,gamarnik-pnas}, which gives a geometric theory of
when planted recovery is blocked for whole classes of algorithms.

\paragraph{What is proved, and what is conjectured.} We work in a planted-spike model
$\hat\mu = (\alpha/\sqrt k)\sigma^\star + g$ that captures the
factor-plus-residual structure documented in \S\ref{sec:diagnostics}. For the
\emph{linear} score $H(\sigma)=\hat\mu^\top\sigma$ we prove a detection upper bound
(\cref{thm:detect}) and a single-sided forbidden-overlap interval (\cref{thm:forbid}).
On this linear model the optimisation problem is in fact solved trivially by top-$k$
thresholding once $a>a_{\mathrm{rec}}$; the theorems are interesting because they
quantify the information-theoretic boundary between random instances and the planted
signal. For the \emph{quadratic} objective $J$ in \eqref{eq:problem-intro}, the classical
interior OGP shape of \cite{gamarnik-zadik-clique} is a conjecture supported by our
empirical evidence: the mixing time of constrained Metropolis blows up sharply between
$(n,k)=(20,5)$ and $(40,10)$ (\S\ref{ssec:p2b}).

\paragraph{Roadmap.}\S\ref{sec:diagnostics} characterises the empirical regime
(sub-Gaussian tails, factor-spike covariance, RIP failure). \S\ref{sec:oracle} establishes
the brute-force oracle, four polynomial-time baselines, and the constrained-Metropolis
sampler whose mixing-time blow-up is our primary empirical OGP evidence.
\S\ref{sec:ogp} proves the linear-model theorems and derives a hardness statement for
algorithms whose iterates are Hamming-stable along an input bridge.
\S\ref{sec:ogp-empirical} verifies the theorems on synthetic data and reports the
analogous (still smooth) overlap distribution for the full quadratic objective on real
$(30,6)$ instances. \S\ref{sec:slh} introduces our hybrid algorithm \textsc{Slh}\ and
proves an exact-recovery proposition under factor-plus-residual covariance.
\S\ref{sec:conclusion} returns to the two questions of the introduction and gives a
direct answer.

% Phase 1: empirical diagnostics
\input{phase1_diagnostics.tex}

% Phase 2: oracle, baselines, MCMC
\input{phase2_oracle_mcmc.tex}

% Phase 3a: OGP theory
\input{phase3a_ogp.tex}

% Phase 3b: empirical OGP experiment
\input{phase3b_empirical.tex}

% Phase 4a: SLH algorithm
\input{phase4_hybrid.tex}

% Conclusion / discussion
\input{conclusion.tex}

\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{markowitz}
H.~Markowitz,
\newblock Portfolio selection,
\newblock {\em Journal of Finance}, 7(1):77-91, 1952.

\bibitem{cramer-original}
H.~Cram\'er,
\newblock Sur un nouveau th\'eor\`eme-limite de la th\'eorie des probabilit\'es,
\newblock {\em Actualit\'es Scientifiques et Industrielles}, 736:5-23, 1938.

\bibitem{vershynin}
R.~Vershynin,
\newblock {\em High-Dimensional Probability: An Introduction with Applications in Data Science},
\newblock Cambridge University Press, 2018.

\bibitem{wainwright}
M.~J.~Wainwright,
\newblock {\em High-Dimensional Statistics: A Non-Asymptotic Viewpoint},
\newblock Cambridge University Press, 2019.

\bibitem{candes-tao-rip}
E.~J.~Cand\`es and T.~Tao,
\newblock Decoding by linear programming,
\newblock {\em IEEE Trans.\ Inform.\ Theory}, 51(12):4203-4215, 2005.

\bibitem{marchenko-pastur}
V.~A.~Marchenko and L.~A.~Pastur,
\newblock Distribution of eigenvalues for some sets of random matrices,
\newblock {\em Mat.\ Sb.}, 72(4):507-536, 1967.

\bibitem{slepian}
D.~Slepian,
\newblock The one-sided barrier problem for Gaussian noise,
\newblock {\em Bell System Tech.\ J.}, 41(2):463-501, 1962.

\bibitem{sudakov-fernique}
X.~Fernique,
\newblock Regularit\'e des trajectoires des fonctions al\'eatoires gaussiennes,
\newblock {\em \'Ecole d'\'Et\'e de Probabilit\'es de Saint-Flour IV-1974}, LNM 480:1-96, 1975.

\bibitem{lindeberg}
J.~W.~Lindeberg,
\newblock Eine neue Herleitung des Exponentialgesetzes in der Wahrscheinlichkeitsrechnung,
\newblock {\em Mathematische Zeitschrift}, 15(1):211-225, 1922.

\bibitem{talagrand-concentration}
M.~Talagrand,
\newblock Concentration of measure and isoperimetric inequalities in product spaces,
\newblock {\em Publ.\ Math.\ IH\'ES}, 81:73-205, 1995.

\bibitem{janson}
S.~Janson,
\newblock Poisson approximation for large deviations,
\newblock {\em Random Structures \& Algorithms}, 1(2):221-229, 1990.

\bibitem{sherrington-kirkpatrick}
D.~Sherrington and S.~Kirkpatrick,
\newblock Solvable model of a spin-glass,
\newblock {\em Physical Review Letters}, 35(26):1792-1796, 1975.

\bibitem{gamarnik-zadik-clique}
D.~Gamarnik and I.~Zadik,
\newblock The landscape of the planted clique problem: Dense subgraphs and the overlap gap property,
\newblock {\em arXiv:1904.07174}, 2019.

\bibitem{gamarnik-pnas}
D.~Gamarnik,
\newblock The overlap gap property: A topological barrier to optimizing over random structures,
\newblock {\em Proceedings of the National Academy of Sciences}, 118(41):e2108492118, 2021.

\bibitem{ben-arous-wein}
G.~Ben~Arous and A.~S.~Wein,
\newblock Free energy wells and overlap gap property in sparse PCA,
\newblock {\em Proc.\ Conference on Learning Theory (COLT)}, pp.~479-482, 2020.

\bibitem{gamarnik-jagannath-wein}
D.~Gamarnik, A.~Jagannath, and A.~S.~Wein,
\newblock Low-degree hardness of random optimization problems,
\newblock {\em Proc.\ FOCS}, pp.~131-140, 2020.

\bibitem{tibshirani-lasso}
R.~Tibshirani,
\newblock Regression shrinkage and selection via the lasso,
\newblock {\em J.~R.~Stat.\ Soc.\ B}, 58(1):267-288, 1996.

\bibitem{ledoit-wolf}
O.~Ledoit and M.~Wolf,
\newblock A well-conditioned estimator for large-dimensional covariance matrices,
\newblock {\em Journal of Multivariate Analysis}, 88(2):365-411, 2004.

\end{thebibliography}

\end{document}
