% Phase 3a: Theoretical analysis: detection threshold and forbidden-overlap interval
% Self-contained section, designed to be \input into the main writeup.

\section{Theory: detection threshold and forbidden-overlap interval}
\label{sec:ogp}

This section gives the project's main rigorous contribution: two theorems on the
\emph{linear} planted-spike score $H(\sigma)=\hat\mu^\top\sigma$ and an algorithmic
hardness corollary for Hamming-stable algorithms. The scope is deliberately limited.

\paragraph{What we prove (linear $H$).}
A first-moment upper bound on a detection threshold $a_\star(\rho)$ (\cref{thm:detect})
and a single-sided forbidden-overlap interval $[\rho, y_a(\rho)]$ for all near-optima
(\cref{thm:forbid}). Both bounds are matched empirically in \S\ref{sec:ogp-empirical}.
On the linear model the optimisation $\arg\max_\sigma H(\sigma)$ is trivially solved by
top-$k$ thresholding of $\hat\mu$; the value of these theorems is informational
(quantifying the boundary between random noise and the planted signal), not algorithmic.
The bounds are first-moment only: we do not prove a matching second-moment lower bound,
so ``sharp'' would be inappropriate.

\paragraph{What we leave open (quadratic $J$).}
The full objective $J(\sigma)=\tfrac{1}{k}\hat\mu^\top\sigma-\tfrac{\gamma}{2k^2}
\sigma^\top\hat\Sigma\sigma$ contains a quadratic spin-glass-like coupling between
subsets that is not captured by the linear model. A classical interior OGP for $J$
\cite{gamarnik-zadik-clique,gamarnik-pnas} is conjecturally responsible for the
mixing-time blow-up observed in \S\ref{ssec:p2b} on $(n,k)=(40,10)$. A rigorous proof
requires a Sherrington-Kirkpatrick-type second-moment computation under
factor-plus-residual covariance \cite{sherrington-kirkpatrick} and is left open.

\paragraph{What the hardness corollary says, and does not say.}
\cref{cor:hardness} excludes a specific class of algorithms (those whose support is
$\eta$-Hamming-stable along a Gaussian input bridge with $\eta < k(y_a-\rho)/2$).
We discuss in \S\ref{ssec:cor-scope} that LASSO and top-eigenvector projection lie in
this class, while greedy forward selection and the constrained-Metropolis chain do not.
We will \emph{not} use this corollary to argue that \textsc{Slh}\ ``circumvents'' an OGP
barrier on real S\&P data: as we explain in \S\ref{sec:slh}, the empirical SNR sits in
the recovery regime (Regime~III) where top-$k$ thresholding alone suffices.

\subsection{Planted-spike portfolio model}

The construction below is the portfolio-selection analogue of the planted models familiar from
the planted-clique problem \cite{gamarnik-zadik-clique} and sparse PCA \cite{ben-arous-wein}.

Fix integers $n,k$ with $k = \rho\, n$ for some density $\rho\in(0,1/2)$. The slice
$\Omega_{n,k} := \{\sigma\in\{0,1\}^n : \sum_{i}\sigma_i = k\}$ encodes cardinality-$k$
portfolios as binary supports. Fix a planted target $\sigma^\star\in\Omega_{n,k}$. We
assume the estimated mean-return vector takes the form
\begin{equation}
\hat\mu \;=\; \frac{\alpha}{\sqrt{k}}\,\sigma^\star \;+\; g, \qquad g\sim\mathcal N(0, I_n),
\label{eq:model}
\end{equation}
with signal-to-noise parameter $\alpha\geq 0$, and we work with the linear
score
\begin{equation}
H(\sigma) \;:=\; \hat\mu^\top\sigma \;=\; \alpha\sqrt{k}\,y(\sigma) + g^\top\sigma,
\qquad y(\sigma) := \frac{\langle\sigma,\sigma^\star\rangle}{k}\in[0,1].
\label{eq:H}
\end{equation}
Conditional on $\sigma$ (with $\sigma^\star$ fixed), $H(\sigma)\sim\mathcal N(\alpha\sqrt
k\,y(\sigma),\, k)$: the variance is independent of $\sigma$ because $\|\sigma\|_2^2=k$
for every $\sigma\in\Omega_{n,k}$.

\paragraph{Why this idealization.} Phase~1 showed that the empirical covariance is a
rank-$r^\star$ factor structure plus a near-isotropic residual; we have folded the
deterministic part of the variance term $\frac{\gamma}{2k^2}\sigma^\top\hat\Sigma\sigma$
into the planted direction $\sigma^\star$ and absorbed the residual into~$g$. The empirical
$\psi_2$-norms (median 3.6) are finite, so~\eqref{eq:model} is a Bernstein-faithful
caricature.

\paragraph{Sub-Gaussian universality (Lindeberg).} The proofs below use only the upper
bound $\overline\Phi(x)\le e^{-x^2/2}$ on the Gaussian survival function at scale
$\sqrt k$. If $g_i$ is instead sub-Gaussian with proxy $\sigma_g^2 = \|g_i\|_{\psi_2}^2$,
the same bound holds with $\sigma_g^2$ in place of $1$, shifting the threshold to
$a_\star(\rho;\sigma_g) = \sigma_g\,(1-\rho)^{-1}\sqrt{2\,h(\rho)}$. The Lindeberg
swapping argument \cite{lindeberg} (see \cite[Ch.~11.1]{vershynin} for a modern treatment)
shows that the conclusions of \cref{thm:detect,thm:forbid} are \emph{universal} in the
noise distribution provided $\mathrm{Var}(g_i)=\sigma_g^2$ and the third absolute moment
$\mathbb E|g_i|^3$ is bounded; the third-moment correction enters the rate as a
lower-order term $O(n^{-1/2})$. The empirical $\psi_2$-norms (median $3.63$, all finite)
imply $\mathbb E|R_{:,i}|^3$ is finite for every stock, so the Lindeberg argument
activates.

\paragraph{Boltzmann measure on the slice.} For $\beta>0$ we use the standard Gibbs measure
$\pi_\beta(\sigma) := \frac{1}{Z(\beta)}\,e^{\beta H(\sigma)}$ with
$Z(\beta) := \sum_{\sigma\in\Omega_{n,k}} e^{\beta H(\sigma)}$. Phase~2b validated a
swap-Metropolis chain whose stationary distribution is exactly $\pi_\beta$ (KS p-values
$\geq 0.42$ across $\beta\in\{0,50,200,1000,5000\}$).

\subsection{First-moment formula and the function $L(y;a,\rho)$}

Define, for a threshold $T\in\mathbb R$ and a planted-overlap value $y\in
\frac{1}{k}\{0,1,\dots,k\}\cap[0,1]$,
\[
A_T(y) \;:=\; \bigl\{\sigma\in\Omega_{n,k} : H(\sigma)\geq T,\ y(\sigma)=y\bigr\}.
\]
The number of $\sigma\in\Omega_{n,k}$ with planted overlap exactly $y$ is
$N_y := \binom{k}{ky}\binom{n-k}{k(1-y)}$.

\begin{lemma}[First-moment formula]\label{lem:firstmoment}
For each admissible $y$, $\mathbb E\bigl[|A_T(y)|\bigr] = N_y\cdot\overline\Phi\!\left(\frac{T-\alpha\sqrt k\,y}{\sqrt k}\right)$,
where $\overline\Phi(x)=\Pr[\mathcal N(0,1)>x]$ is the Gaussian survival function.
\end{lemma}
\begin{proof}
By linearity, $\mathbb E|A_T(y)| = \sum_{\sigma:y(\sigma)=y}\Pr[H(\sigma)\geq T]$. Each
summand equals $\overline\Phi((T-\alpha\sqrt k y)/\sqrt k)$ by~\eqref{eq:H}, and there are
$N_y$ terms.
\end{proof}

Take $\rho$ fixed, $\alpha=a\sqrt n$ with $a>0$ a constant, and threshold $T = a\sqrt{nk}$
(the expected score of the planted). Stirling and the standard Gaussian tail bound
$\overline\Phi(x)\leq e^{-x^2/2}$ give (a Cram\'er-style large-deviation rate function \cite{cramer-original} in disguise)
\begin{equation}
\frac{1}{n}\log\mathbb E|A_T(y)| \;\leq\; L(y;a,\rho) + o(1),\qquad
\boxed{\ L(y;\,a,\rho) := s_\rho(y) - \tfrac{a^2}{2}(1-y)^2,\ }
\label{eq:Ldef}
\end{equation}
where the slice entropy is
\begin{equation}
s_\rho(y) \;:=\; \rho\,h(y) + (1-\rho)\,h\!\left(\tfrac{\rho(1-y)}{1-\rho}\right),
\qquad h(p) := -p\log p - (1-p)\log(1-p).
\label{eq:srho}
\end{equation}
The function $L(\,\cdot\,;a,\rho)$ is concave on $[0,1]$ with $L(1)=0$ (the planted is the
unique configuration at $y=1$, contributing entropy zero), and its sign on the interior
governs whether $T$-near-optima exist at planted-overlap~$y$.

\subsection{Detection upper bound (Theorem~\ref{thm:detect})}

\begin{theorem}[First-moment detection upper bound]\label{thm:detect}
Let $\rho\in(0,1/2)$ and define
\begin{equation}\label{eq:adetect}
a_\star(\rho) \;:=\; \frac{1}{1-\rho}\,\sqrt{2\,h(\rho)}.
\end{equation}
For every $a > a_\star(\rho)$ there exist $\delta>0$ and $c=c(a,\rho,\delta)>0$ such that,
under the planted model~\eqref{eq:model} with $\alpha=a\sqrt n$ and threshold $T=a\sqrt{nk}$,
\[
\mathbb P\bigl[\,\exists\,\sigma\in\Omega_{n,k}\text{ with } y(\sigma)\le \rho+\delta
\text{ and } H(\sigma)\ge T\,\bigr]\;\le\; e^{-c n}.
\]
\end{theorem}
\begin{proof}
At $y = \rho$, $s_\rho(\rho) = h(\rho)$, so $L(\rho;a,\rho) = h(\rho) - \tfrac{a^2}{2}(1-\rho)^2 < 0$
iff $a > a_\star(\rho)$. By continuity of $y\mapsto L(y;a,\rho)$, pick $\delta>0$ with
$L(y;a,\rho)\le -c'<0$ uniformly on $y\in[0,\rho+\delta]$. The admissible values of
$y(\sigma)$ for $\sigma\in\Omega_{n,k}$ form a finite lattice
$\{j/k : 0\le j\le k\}$, so the number of admissible $y$ in $[0,\rho+\delta]$ is at most
$\lceil(\rho+\delta)k\rceil + 1 \le k + 1 = O(n)$. Union-bounding Markov's inequality with
the rate bound~\eqref{eq:Ldef} over these values gives
\[
\Pr[\exists \sigma\text{ with } y(\sigma)\in[0,\rho+\delta], H(\sigma)\ge T]
  \;\le\; (k+1)\, e^{-c' n}\,(1+o(1)) \;\le\; e^{-(c'/2)\,n},
\]
where the polynomial prefactor $k+1$ is absorbed into the exponent for $n$ large enough.
\end{proof}

\paragraph{Numerical example.} With $\rho=0.20$, the formula gives $a_\star(0.20)=(1-0.20)^{-1}\sqrt{2\,h(0.20)}\approx 1.250$.
Below this SNR the planted spike is statistically undetectable in the sense that the
expected number of random subsets achieving the planted score is exponentially large;
above it, it is exponentially small.

\paragraph{No matching lower bound.} We prove the upper bound on the detection threshold
via the first moment only. A matching lower bound (showing that for $a<a_\star(\rho)$
some specific algorithm succeeds, or that some specific test fails) would require a
second-moment computation $\mathbb E|A_T(y)|^2/(\mathbb E|A_T(y)|)^2 = 1+o(1)$ on the
slice. For the planted-clique problem this was carried out by
\cite{gamarnik-zadik-clique}; we conjecture that the same argument applies here and
that $a_\star(\rho)$ is the sharp threshold, but we do not prove it.

\subsection{Forbidden-overlap interval (Theorem~\ref{thm:forbid})}

The detection theorem only excludes near-optima from a neighborhood of $y=\rho$. The
following statement extends the forbidden region: above detection, no near-optimum has
planted-overlap below an explicit threshold $y_a\in(\rho,1)$.

\begin{theorem}[Forbidden-overlap interval]\label{thm:forbid}
Let $\rho\in(0,1/2)$, $\alpha=a\sqrt n$ with $a > a_\star(\rho)$, and $T = a\sqrt{nk}$. Define
\begin{equation}\label{eq:ya}
y_a(\rho) \;:=\; \sup\bigl\{y\in[\rho,1) : L(y;a,\rho) \leq 0\bigr\}.
\end{equation}
Then $y_a(\rho)\in(\rho, 1)$ is well-defined, and for every sufficiently small $\delta>0$
there exists $c=c(a,\rho,\delta)>0$ with
\begin{equation}\label{eq:forbid}
\Pr\bigl[\,\exists\,\sigma\in\Omega_{n,k} : \rho+\delta \le y(\sigma) \le y_a(\rho) - \delta
\text{ and } H(\sigma) \ge T\,\bigr] \;\le\; e^{-c n}.
\end{equation}
Equivalently, every $T$-near-optimum has planted-overlap either below $\rho+\delta$
(excluded by \cref{thm:detect}) or above $y_a(\rho)-\delta$.
\end{theorem}
\begin{proof}
The function $y\mapsto L(y;a,\rho)$ is continuous and concave on $[0,1]$ with
$L(\rho;a,\rho)<0$ (since $a > a_\star(\rho)$), $L(1;a,\rho) = 0$, and
$\partial_y L(y;a,\rho)\big|_{y\to 1^-}\to-\infty$ (the entropy term $h(\rho(1-y)/(1-\rho))$
has slope $-\infty$ as $y\to 1$). Hence $L$ has exactly one zero
$y_a\in(\rho,1)$ in the open interval $(\rho,1)$, and $L<0$ on $(\rho, y_a)$.

For the probability bound, let $c' := -\sup_{y\in[\rho+\delta, y_a-\delta]} L(y;a,\rho) > 0$
(strict positivity on a compact subinterval). The admissible discrete values of
$y(\sigma)$ form the lattice $\{j/k : 0\le j\le k\}$, so the number of admissible $y$ in
$[\rho+\delta,y_a-\delta]$ is at most $k+1\le n+1$. By Markov's inequality combined with
\eqref{eq:Ldef},
\begin{equation}\label{eq:union}
\Pr[\exists\sigma\text{ in the forbidden band, } H(\sigma)\ge T]
  \;\le\; \sum_{y\in\text{band}} \mathbb E|A_T(y)|
  \;\le\; (n+1)\,e^{-c' n + o(n)}
  \;\le\; e^{-(c'/2)\,n}
\end{equation}
for $n$ large enough. (The exponential bound dominates the linear prefactor; one can
sharpen the prefactor from $n$ to $1$ via Talagrand-style concentration of $|A_T(y)|$
around its mean \cite{talagrand-concentration}, or Janson-style large-deviation bounds
\cite{janson}, but the present bound already gives the exponential rate that matters.)
\end{proof}

\paragraph{Numerical example.} For $\rho=0.20$ and $a = 2.0$ (well above the detection
threshold $a_\star(0.20)\approx 1.25$), numerical root-finding in the companion notebook
\texttt{04a\_ogp\_theory.ipynb} gives $y_a\approx 0.55$. So no near-optimum exists with
planted overlap in $[0.21, 0.54]$; all near-optima have $y\geq 0.55$ or $y\leq 0.20$, the
latter excluded by Theorem~\ref{thm:detect}.

\subsection{Algorithmic consequence: a constrained hardness corollary}\label{ssec:cor-scope}

The forbidden-overlap interval \eqref{eq:forbid} translates into a hardness statement for
algorithms whose iterates are Hamming-stable along an input-noise interpolation. The class
is narrow; we are careful below to delineate exactly which algorithms it captures and
which it does \emph{not}.

The class of $\eta$-Hamming-stable algorithms below is closely related (but not identical)
to the low-degree polynomial estimator class for which a comprehensive hardness theory exists
\cite{gamarnik-jagannath-wein}.

\begin{definition}[$\eta$-Hamming-stable along a Gaussian bridge]\label{def:stable}
Let $g_0,g_1\in\mathbb R^n$ be two i.i.d.\ realizations of the noise and define the
Gaussian bridge $g_s := \sqrt{1-s}\,g_0 + \sqrt{s}\,g_1$, $s\in[0,1]$, with
$\hat\mu_s = (\alpha/\sqrt k)\,\sigma^\star + g_s$. An algorithm
$\mathcal A:\mathbb R^n\to\Omega_{n,k}$ is \emph{$\eta$-Hamming-stable along the bridge}
if the discrete-time evolution $s\mapsto\mathcal A(\hat\mu_s)$ on the lattice
$s\in\{0,\tfrac1n,\dots,1\}$ has total Hamming displacement
$\sum_{s} d_H(\mathcal A(\hat\mu_{s_-}),\mathcal A(\hat\mu_s)) \le \eta\,k$
with probability $1-o(1)$.
\end{definition}

\begin{corollary}[Hardness for path-stable algorithms]\label{cor:hardness}
Under the hypotheses of Theorem~\ref{thm:forbid}, fix $\eta < (y_a(\rho)-\rho)/2$.
Suppose the bridge endpoint $\hat\mu_0$ is such that $\mathcal A(\hat\mu_0)$ has overlap
$y(\mathcal A(\hat\mu_0))\le \rho + o(1)$ (e.g.\ $\mathcal A$ is invariant in distribution
under permutations of $[n]$ and the planted target is sampled uniformly). Then for every
$\eta$-Hamming-stable algorithm $\mathcal A$ satisfying $\mathcal A(\hat\mu_1)=\sigma^\star$,
\[
\Pr\bigl[\,H(\mathcal A(\hat\mu_s))\ge T \ \text{for every}\ s\in\{0,\tfrac1n,\dots,1\}\bigr]
\;\le\; o(1) + e^{-c n}.
\]
\end{corollary}
\begin{proof}[Sketch]
By stability, the total Hamming displacement along the lattice is $\le \eta k$; in particular
the largest single-step jump is $\le \eta k$, so $|y(\mathcal A(\hat\mu_{s+1/n})) -
y(\mathcal A(\hat\mu_s))| \le \eta < (y_a-\rho)/2$. Since
$y(\mathcal A(\hat\mu_0))\le\rho$ and (under recovery) $y(\mathcal A(\hat\mu_1))=1$, the
discrete intermediate-value theorem produces some $s^\star\in[0,1]$ with
$y(\mathcal A(\hat\mu_{s^\star}))\in(\rho+\delta, y_a(\rho)-\delta)$. \cref{thm:forbid}
then forces $H(\mathcal A(\hat\mu_{s^\star}))<T$ with probability $1-e^{-cn}$.
\end{proof}

\paragraph{Scope of the corollary.}
The corollary excludes a specific class of algorithms; it is silent about others. We are
careful in what we claim:
\begin{itemize}\itemsep=2pt
\item \textbf{LASSO-Markowitz} (\S\ref{sec:lasso}) is a Lipschitz function of $\hat\mu$;
rounding to top-$k$ support introduces only $O(1)$ Hamming jumps along the bridge, so
LASSO is $\eta$-stable for $\eta = o(1)$. The corollary then predicts that for
$a>a_\star(\rho)$, LASSO cannot maintain near-optimality along the input path. Phase~2a
reports LASSO gaps of $4$ to $11$\,bp uniformly across $(n,k)$, consistent with the
prediction but \emph{not} a proof of it (Phase~2a is the empirical $J$-objective, not the
linear $H$-objective the corollary controls).

\item \textbf{Top-eigenvector projection} is similarly $\eta$-stable outside a
Lebesgue-null set and falls in the corollary's class.

\item \textbf{Greedy forward selection} is \emph{not} $\eta$-Hamming-stable: a single
infinitesimal perturbation of $\hat\mu$ can swap two indices near the rank-$k/k{+}1$
boundary and trigger a combinatorial cascade. Phase~2a observes greedy tracks the oracle
closely, consistent with the corollary not applying.

\item \textbf{Top-$k$ thresholding of $\hat\mu$} (the seed stage of our \textsc{Slh}\ algorithm)
is $\eta$-stable with $\eta=O(1)$ outside Lebesgue-null ties: a small Gaussian
perturbation moves at most $O(1)$ coordinates across the $k$-th-largest threshold per
sub-interval of the bridge. Hence top-$k$ thresholding \emph{is in the corollary's
class}, but the contradiction with the corollary is only triggered when $a < a_{\mathrm{rec}}$
(the recovery threshold of \cref{prop:slh-recovery}); for $a > a_{\mathrm{rec}}$ the
required ``$\mathcal A(\hat\mu_0)$ has $y\le\rho$'' hypothesis fails on the bridge, and the
corollary is vacuous. This is the key observation: on real S\&P data we operate in the
recovery regime, not the OGP-active regime, so the corollary does not bind.

\item \textbf{The constrained-Metropolis chain} of \S\ref{ssec:p2b} is not stable under
\cref{def:stable}: it is the very dynamics whose mixing-time blow-up at $(40,10)$
provides our primary empirical evidence \emph{that} a geometric barrier is present on
the (full quadratic) objective $J$.
\end{itemize}

\paragraph{What the corollary does not establish.} The corollary is stated for the linear
score $H$, not the empirical objective $J$. It does \emph{not}, by itself, certify that
any specific algorithm fails on real S\&P data. The empirical observation that LASSO and
top-eigenvector underperform the oracle (Phase~2a) is consistent with the corollary, not
proof of it.

\subsection{Discussion: classical OGP and what is conjectural}
\label{ssec:ogp-discussion}

Theorems~\ref{thm:detect} and \ref{thm:forbid} produce a forbidden-overlap interval that is
\emph{single-sided}: the forbidden region is $[\rho+\delta, y_a(\rho)-\delta]$, a single
interval terminating before the planted point. The classical OGP shape of \cite{gamarnik-zadik-clique,gamarnik-pnas}
is a forbidden interval $(y_\ell, y_h)\subsetneq(\rho,1)$ that is \emph{interior}, separating
two non-trivial allowed components. We close with three remarks.

\begin{enumerate}
\item \textbf{Why the simple model has no interior forbidden gap.} The threshold function
$a_\star(y;\rho) := (1-y)^{-1}\sqrt{2\, s_\rho(y)}$ is monotonically increasing on $(\rho,1)$
(numerically verified in the companion notebook for $\rho\in[0.05, 0.40]$). Hence the
sub-level set $\{y : a_\star(y;\rho)<a\}$ is a single connected interval $[\rho, y_a(\rho)]$
for every $a>a_\star(\rho;\rho)$, ruling out a classical OGP forbidden interval from
first-moment alone.

\item \textbf{Where the classical OGP comes from in the full model.} The empirical
Hamiltonian $J(\sigma) = \tfrac{1}{k}\hat\mu^\top\sigma - \tfrac{\gamma}{2k^2}\sigma^\top
\hat\Sigma\sigma$ contains a quadratic form that is \emph{not} captured
by~\eqref{eq:model}. Phase~1 showed $\hat\Sigma$ has $32$ outlier eigenvalues, so the
quadratic form is a low-rank deterministic spike plus an isotropic random part. The
isotropic random part contributes a Sherrington-Kirkpatrick-like coupling between subsets
that, by the standard second-moment argument
\cite{sherrington-kirkpatrick,gamarnik-pnas}, is expected to produce an
interior forbidden interval at intermediate inverse temperature.

\item \textbf{Empirical verification.} The forbidden-overlap interval in
Theorem~\ref{thm:forbid} predicts that MCMC samples drawn from $\pi_\beta$ at high $\beta$
should have planted-overlap distribution concentrated either near $y=\rho$ or near $y=1$,
with no mass at intermediate overlaps. Phase~3b runs this experiment on synthetic data
matched to the empirical SNR and on the real S\&P~500 panel, and reports the bimodality
both for the linear model (where Theorem~\ref{thm:forbid} predicts it directly) and for
the full quadratic objective (where the additional interior forbidden gap is conjectural
but expected from~(2) above).
\end{enumerate}

The detection threshold of Theorem~\ref{thm:detect} and the forbidden-overlap interval
of Theorem~\ref{thm:forbid} are the rigorous theoretical contributions of the project.
The companion notebook \texttt{04a\_ogp\_theory.ipynb} verifies the threshold numerically
on a $(\rho, a)$ grid and produces Figures~\ref{fig:Lcurve},
\ref{fig:detection-curve}, and~\ref{fig:phase} of \S\ref{sec:ogp-empirical}.
